Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-15 talk/pepm2011/paper.tex
enhance example
Rewording.
incorporate first round of comments by stephan

\maketitle
\begin{abstract}
The performance of many dynamic language implementations suffers from
high allocation rates and runtime type checks.  This makes dynamic
languages less applicable to purely algorithmic problems, despite their
growing popularity.  In this paper, we present a simple optimization
based on online partial evaluation to remove object allocations and
runtime type checks in the context of a tracing JIT.  We evaluate the
optimization using a Python VM and find that it gives good results for
all our (real-life) benchmarks.
\footnote{This research is partially supported by the BMBF funded project PyJIT (nr. 01QE0913B;
Eureka Eurostars).}
\end{abstract}

@@

The goal of a just-in-time (JIT) compiler for a dynamic language is obviously to
improve the speed of the language over an implementation of the language that
uses interpretation. The first goal of a JIT is therefore to remove the
interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
overhead of the interpreter's data structures, such as operand stack etc. The
second important problem that any JIT for a dynamic language needs to solve is
how to deal with the overhead of boxing primitive types and of type
dispatching. Those are problems that are usually not present or at least less
severe in statically typed languages.

Boxing of primitive types is necessary because dynamic languages need to be able to handle
all objects, even integers, floats, booleans etc. in the same way as user-defined
instances. Thus those primitive types are usually \emph{boxed}, i.e. a small
heap-structure is allocated for them, that contains the actual value. Boxing
primitive types can be very costly, because a lot of common operations,
particularly all arithmetic operations, have to produce a new box, in addition
to the actual computation they do. Because the boxes are allocated on the heap,
producing a lot of them puts pressure on the garbage collector.

Type dispatching is the process of finding the concrete implementation that is
applicable to the objects at hand when doing a generic operation on them. An
example would be the addition of two objects: The addition needs to check what
the concrete objects that should be added are, and choose the implementation
that is fitting for them. Type dispatching is a very common operation in
modern\footnote{For languages in the LISP family, basic arithmetic operations
are typically not overloaded; even in Smalltalk, type dispatching is much
simpler than in Python or JavaScript.}
dynamic languages because no types are known at compile time, so all operations
need it.

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot spots into linear execution traces. Working on
traces is the central idea of a tracing JIT. Those traces are optimized and
turned into machine code.

One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of the interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many algorithms that are usually
hard in a compiler, such as register allocation.

The usage of a tracing JIT can remove the overhead of bytecode dispatch and that
of the interpreter data structures. In this paper we want to present a new
optimization that can be added to a tracing JIT that further removes some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. Our experimental platform is the PyPy project, which is an
environment for implementing dynamic programming languages. PyPy and tracing
JITs are described in more detail in Section~\ref{sec:Background}.
Section~\ref{sec:lifetimes} analyzes the problem to be solved more closely.

The core of our trace optimization technique can be
viewed as partial evaluation: the partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and makes some
objects that are allocated in the trace \emph{static,} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}; a more formal description is
given in Section~\ref{sec:formal}. The introduced
techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
interpreter as a case study.

The contributions of this paper are:

\begin{enumerate}
    \item An efficient and effective algorithm for removing object allocations in a tracing JIT.
    \item A characterization of this algorithm as partial evaluation.
    \item A rigorous evaluation of this algorithm.
\end{enumerate}

@@

The work described in this paper was done in the context of the PyPy project\footnote{\texttt{http://pypy.org}}
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. 
When implementing a language with PyPy one writes an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. During translation to C, many low-level
aspects of the final VM, such as object layout, garbage collection and memory
model, are woven into the generated code. Therefore the interpreter itself can
remain at a relatively high level of abstraction.

@@

The feature that makes PyPy more than a compiler with a runtime system is its
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a tracing just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer using a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Many optimizations can benefit all languages implemented as an interpreter in RPython.

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code.  Examples of
JIT backends that are implemented are the one for Intel x86 and x86-64 and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
mostly mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter tries to record all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.

This trace of operations is then the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimization
and machine code generation are simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

Since the trace corresponds to one concrete execution of a loop,
the code generated from it is only one possible path through it.
To make sure that the trace is maintaining the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an \lstinline{if} statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the \lstinline{if} statement is the same as
during tracing, because if
it isn't, the rest of the trace is not valid. \cfbolz{The "if" shouldn't be bold}

@@

Using these classes to implement arithmetic shows the basic problem that a
dynamic language implementation has. All the numbers are instances of either
\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
which puts pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

@@

The loop in \lstinline{f} iterates \lstinline{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \lstinline{is_positive} and even two for each
call to \lstinline{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects do not survive for very long.
The actual computation that is performed by \lstinline{f} is simply a sequence of
float or integer additions.

@@

In the example from last section, the following operations in the upper half
of Fig.~\ref{fig:unopt-trace} produce two
static objects, and can be completely removed from the optimized trace:

@@

The subsequent operations in Fig.~\ref{fig:unopt-trace},
 which use $p_{5}$ and $p_{6}$, could then be
optimized using that knowledge:

@@

The rest of the trace from Fig.~\ref{fig:unopt-trace} is optimized similarly.

@@

This is what happens at the end of the trace in Fig.~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively). This means that instead of a simple jump,
the following operations are emitted:

@@

After the lifting the static heap is the empty set, because both static objects
were lifted.

>>>>

\maketitle
\begin{abstract}
The performance of many dynamic language implementations suffers from
high allocation rates and runtime type checks.  This makes dynamic
languages less applicable to purely algorithmic problems, despite their
growing popularity.  In this paper we present a simple compiler optimization
based on online partial evaluation to remove object allocations and
runtime type checks in the context of a tracing JIT.  We evaluate the
optimization using a Python VM and find that it gives good results for
all our (real-life) benchmarks.
\footnote{This research is partially supported by the BMBF funded project PyJIT (nr. 01QE0913B;
Eureka Eurostars).}
\end{abstract}

@@

The objective of a just-in-time (JIT) compiler for a dynamic language is to
improve the speed of the language over an implementation of the language that
uses interpretation. The first goal of a JIT is therefore to remove the
interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
overhead of the interpreter's data structures, such as operand stack etc. The
second important problem that any JIT for a dynamic language needs to solve is
how to deal with the overhead of boxing primitive types and of type
dispatching. Those are problems that are usually not present or at least less
severe in statically typed languages.

Boxing of primitive types is necessary because dynamic languages need to be able to handle
all objects, even integers, floats, booleans etc. in the same way as user-defined
instances. Thus those primitive types are usually \emph{boxed}, \ie a small
heap-structure is allocated for them that contains the actual value. Boxing
primitive types can be very costly, because a lot of common operations,
particularly all arithmetic operations, have to produce new boxes, in addition
to the actual computation they do. Because the boxes are allocated on the heap,
producing many of them puts pressure on the garbage collector.

Type dispatching is the process of finding the concrete implementation that is
applicable to the objects at hand when performing a generic operation on them. An
example would be the addition of two objects: For addition the types of the
concrete objects need to be checked and the suiting implementation chosen.
Type dispatching is a very common operation in
modern\footnote{For languages in the LISP family, basic arithmetic operations
are typically not overloaded; even in Smalltalk, type dispatching is much
simpler than in Python or JavaScript.}
dynamic languages because no types are known at compile time. Therefore all
operations need it.

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot spots into \emph{linear execution traces}. Those
traces are optimized and turned into machine code.

One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an existing interpreter, reusing a lot of
the interpreter's infrastructure. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many of the hard algorithms in
a compiler, such as register allocation.

The use of a tracing JIT can remove the overhead of bytecode dispatch and that
of the interpreter data structures. In this paper we want to present a new
optimization that can be added to a tracing JIT that further removes some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. Our experimental platform is the PyPy project, which is an
environment for implementing dynamic programming languages. PyPy and tracing
JITs are described in more detail in Section~\ref{sec:Background}.
Section~\ref{sec:lifetimes} analyzes the problem to be solved more closely.

The core of our trace optimization technique can be
viewed as partial evaluation: the partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and makes some
objects that are allocated in the trace \emph{static,} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}; a more formal description is
given in Section~\ref{sec:formal}. The introduced
techniques are evaluated in Section~\ref{sec:Evaluation} using PyPy's Python
interpreter.

The contributions made by this paper are:

\begin{enumerate}
    \item A description of an efficient and effective algorithm for removing
          object allocations in a tracing JIT.
    \item A characterization of this algorithm as partial evaluation.
    \item Performance benchmarks for this algorithm.
\end{enumerate}

@@

The work described in this paper was done in the context of the PyPy project\footnote{\texttt{http://pypy.org}}
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. 
When implementing a language with PyPy one writes an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on C level. During translation to C, many low-level
aspects of the final VM, such as object layout, garbage collection and memory
model, are woven into the generated code. Therefore the interpreter itself can
remain at a relatively high level of abstraction.

@@

The feature that makes PyPy more than a compiler with a runtime system is its
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a tracing just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer using a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Optimizations can be shared between different languages implemented with PyPy.

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code.  Examples of
JIT backends that are implemented are those for Intel x86 and x86-64 and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the commonly executed, \ie \emph{hot} paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
mostly mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter tries to record all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.
This trace of operations subsequently forms the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimization
and machine code generation are simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

Since the trace corresponds to one concrete execution of a loop,
the code generated from it is only one possible path through the loop.
To make sure that the trace maintains the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an if-statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the if-statement is the same as
during tracing, because if
it isn't, the rest of the trace would not be valid.

@@

Using these classes to implement arithmetic shows the basic problem of a
dynamic language implementation. All the numbers are instances of either
\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
putthing pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

@@

The loop in \lstinline{f} iterates \lstinline{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \lstinline{is_positive} and even two for each
call to \lstinline{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects are short-lived.
The actual computation that is performed by \lstinline{f} is simply a sequence of
float or integer additions.

@@

In the example from last section, the following operations in the upper half
of Figure~\ref{fig:unopt-trace} produce two
static objects, and can be completely removed from the optimized trace:

@@

The subsequent operations in Figure~\ref{fig:unopt-trace},
 which use $p_{5}$ and $p_{6}$, could then be
optimized using that knowledge:

@@

The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized similarly.

@@

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively). This means that instead of a simple jump,
the following operations are emitted:

@@

After the lifting the static heap is the empty set, because both static objects
were lifted.
If we had lifted $w^*$ instead of $v^*$, then the following operations would
have been produced:

\begin{lstlisting}[mathescape,xleftmargin=20pt]
$w^*$ = new($T_2$)
set($w^*$, $L$, $u^*$)
set($w^*$, $R$, $u^*$)
\end{lstlisting}

In this case, the static heap afterwards would be
$$\{v^* \mapsto (T_1, w^*, v^*)\}$$.
